Drought indices of water demand and supply, soil moisture, vegetation, and their impact on LULCC in continental Chile

### 2024-01-22 article/paper_drought_Chile2023.qmd
actualizar repo

title: Drought indices of water demand and supply, soil moisture, vegetation, and its impact on LULCC in continental Chile

@@

  Human-induced greenhouse gas emissions have increased the frequency and/or intensity of weather and climate extremes. Central Chile has been affected by a persistent drought which is impacting the hydrological system and vegetation development. The region has been the focus of research studies due to the diminishing water supply, this persistent period of water scarcity has been defined as a “mega drought". Nevertheless, our results evidence that the water deficit has expanded beyond. Our goal is to analyze the impact of drought, measured by drought indices of water supply/demand and vegetation status, in the LULCC (land use land cover change) over continental Chile.  For the analysis, continental Chile was divided into five zones according to a latitudinal gradient: “Norte Grande”,  “Norte Chico”,  “Zona Central”, “Zona Sur”, and “Zona Austral”.  We used monthly climatic re-analysis variables for precipitation, temperature and soil moisture for 1981-2023 from ERA5-Land (ERA5L); and MODIS (Moderate-Resolution Imaging Spectroradiometer) product MCD12Q1 for land cover for 2001-2021, and the NDVI vegetation index from product MOD13A2 collection 6.1 for 2000-2023, both from collection 6.1. We estimated atmospheric evaporative demand (AED) by combining the Hargreaves-Samani equation with the ERA5L temperature. We derived the drought indices SPI (Standardized Precipitation Index), SPEI (Standardized Precipitation Evapotranspiration Index), EDDI (Evaporative Demand Drought Index), SSI (standardized anomaly of cumulative soil moisture), and the zcNDVI (standardized anomaly of cumulative NDVI). These indices were calculated for time scales of 1, 3, 6, 12, 24, and 36 months, except for zcNDVI (1, 3, and 6 months). We analyzed the temporal correlation of SPI, SPEI, EDDI, and SSI with zcNDVI to have insights into the impact of water supply and demand on vegetation.   Our results showed that LULCC had an increasing trend of 412 [km2yr−1] of forest expansion in the “Zona Sur”, together with a  decreasing trend of 24 [km2yr−1] of cropland contraction in the “Zona Central” meanwhile the “Zona Sur” showed an increase of 31 [km2yr−1], and a contraction of 80 [km2yr−1] of bare soil in the “Zona Austral”. The EDDI was the less correlated index for the five macro zones and the five types of land cover, showing that the temperature in Chile has little impact on vegetation. Higher r-squared values, between 0.5 and 0.8, were obtained at “Norte Chico” and “Zona Central” for the land cover types of savanna, shrubland, grassland, and croplands for the indices SPEI and SSI at time scales of 12 and 24 months. The forest type reaches a r-squared of ~0.5 for SSI of 12 months. The results indicate that the “Norte Chico” and “Zona Central” are the most sensitive regions to water supply deficits longer than a year, potentially explained by a low capacity of water storage in those zones that should be further investigated.

@@

@@

The sixth assessment report (AR6) of the IPCC [@IPCC2023] indicates that human-induced greenhouse gas emissions have increased the frequency and/or intensity of some weather and climate extremes, and the evidence has been strengthened since AR5 [@IPCC2013]. There is high confidence that increasing global warming can expand the land area affected by increasing drought frequency and severity [@IPCCCH112021]. Furthermore, drought increases tree mortality and triggers changes in land cover and, consequently, land use, thus impacting ecosystems [@Crausbay2017]. Nevertheless, there is a lack of understanding of how the alteration in water supply and demand is affecting land cover transformations.

Precipitation is the primary driver of drought and is intensified by temperature [@Luo2017]. Drought impacts soil moisture, hydrological regimes, and vegetation productivity. Initially, drought was commonly classified as meteorological, hydrological, and agricultural [@Wilhite1985]. Lately, @Loon2016 and @AghaKouchak2021 have given an updated definition of drought for the Anthropocene, suggesting that it should be considered the feedback of humans' decisions and activities that drives the anthropogenic drought. Even though it has been argued that those definitions do not fully address the ecological dimensions of drought. @Crausbay2017 proposed the ecological drought definition as *"an episodic deficit in water availability that drives ecosystems beyond thresholds of vulnerability, impacts ecosystem services, and triggers feedback in natural and/or human systems"*. Moreover, many ecological studies have misinterpreted how to characterize drought, for example, sometimes considering "dry" conditions as "drought" [@Slette2019]. On the other hand, the AR6 [@IPCC2023] states that even if global warming is stabilized at 1.5°–2°C, many parts of the world will be impacted by more severe agricultural and ecological droughts. Then, there is a challenge in conducting drought research, especially to evaluate its impact on ecosystems.

@@

For the spatiotemporal assessment of drought impact (i.e., by water supply and demand) on land cover changes, we need climatic realiable variables such as precipitation, temperature, soil moisture, land cover, and vegetation status. For developing countries like Chile, the weather networks present several disadvantages, such as gaps, a short history, and low-quality data. Reanalysis data, as the ERA5-Land (ERA5L) [@MunozSabater2021] provides hourly climatic information (precipitation, temperature, and soil moisture) without gaps since 1950 with global extension. ERA5L has already been used for drought assessment using the Standardized Precipitation-Evapotranspiration Index (SPEI) [@Nouri2023] and for flash drought [@Wang2023] by analyzing soil mositure and evapotranspiration. On the other hand, satellite remote sensing [@West2019; @AghaKouchak2015] is the primary method to evaluate how drought impacts vegetation dynamics. Vegetation drought indices (VDI) are commonly used as proxies of productivity [@Paruelo2016; @Schucknecht2017], which can be derived from the MODIS (Moderate-Resolution Imaging Spectroradiometer).  Besides, land use and land cover (LULC) change can be driven by drought [@Tran2019; @Akinyemi2021]. To analyze these changes, multiple LULC products exist [@Grekousis2015], one of those that provides time series since 2001 is the MCD12Q1 [@Friedl2019] from MODIS. The variation in water supply and demand is finally reflected in the total water storage (TWS). The TWS can be retrieved by the Gravity Recovery and Climate Experiment (GRACE), which allows analyzing water availability changes at coarse resolution [@Ahmed2014; @Ma2017]. We can use climatic reanalysis (ERA5L) and vegetation data (MODIS) to derive drought indices of supply (i.e., precipitation) and demand (i.e., temperature) and thus evaluate the impact of drought on LULC changes. Further, the TWS can be assessed with regard to the changes in water supply and demand to gain insight into the impact on water storage.

@@

In this research, we present the raster dataset DDS4Chl, which provides climate variables and drought indices of water demand and supply and vegetation productivity for continental Chile since 1981 at a monthly frequency. Those were gathered from the earth observation products ERA5L and MODIS. Then, we used DDS4Chl to analyze the impact of drought on different types of land cover classes in continental Chile. The specific objectives of the study are: i) to analyze the trend on multi-scalar drought indices for water demand and supply, soil moisture, and vegetation productivity for 1981–2023/2001–2023; ii) to evaluate LULC change for 2001–2021 and its relation to drought indices; iii) to analyze the relationship of a proxy of vegetation productivity (zcNDVI) with drought indices of water demand and supply and soil moisture; and iv) to assess if the observed changes in the drought indices are linked to TWS.

@@

Continetal Chile has a diverse climate condition from north to south and east to west [@Aceituno2021] (@fig-studyArea), which determines its great ecosystem diversity (@fig-LCprop). The Andes Mountains are a main factor in latitudinal variation [@Garreaud2009]. To describe the climate and ecosystem of Chile, we use the Koppen-Geiger release by @Beck2023 and the landcover type persistance of 80% of times for 2001–2022, from the IGBP classification scheme [@Friedl2019] (see @sec-methods_lulc). "Norte Grande" and "Norte Chico" predominate in an arid desert climate with hot (Bwh) and cold (Bwk) temperatures. At the south of "Norte Chico," the climate changes to an arid steppe with cold temperatures (Bsk). Mainly, the land is barren, with a minor surface of vegetation types such as shrubland and grassland. In the zones "Centro" and the north half of "Sur," the main climate is Mediterranean, with warmer to hot summers (Csa and Csb). In "Centro," there is a major proportion (~50%) of Chilean matorral (shrubland and savanna, [@Fuentes2021]), followed by grassland (~16%), forest (8%), and croplands (5%). The south part of "Sur" and the north part of "Austral" are dominated by an Oceanic climate (Cfb). Those zones are high in forest and grassland. The southern part of the country has a tundra climate, and in Patagonia, it is a cold semi-arid area with an extended surface of grassland, forest, and, to a lesser extent, savanna.

@@

```{=latex}
\begin{table}[!ht]
\caption{Description of the earth observation data used }
\label{tab-desEOD}
\small
\centering
\begin{tabular}{p{0.13\textwidth}cp{0.3\textwidth}p{0.095\textwidth}ccc}
\hline
\multirow{1}{*}{\centering Product} & Sub-product & Variable & Spatial Resolution  & Period & Units & Short Name \\ 
\hline
\multirow{4}{*}{ERA5L} & ~ & Precipitation & \multirow{4}{*}{~0.1°} & \multirow{4}{*}{1981-2023} & mm & P \\ 
         &  & Maximum temperature & ~ & & $°C$ & $T_{max}$ \\ 
         &  & Minimum temperature & ~ & & $°C$ & $T_{min}$ \\ 
         &  & Volumetric Soil Water Content at 1m & ~ & & $m3/m3$ & SM \\ 
ERA5L* & & Atmospheric Evaporative Demand & 0.1° & 1981-2023 & mm & AED \\
        \multirow{2}{*}{MODIS} & MOD13A3.061 & Normalized Difference Vegetation Index & \multirow{2}{*}{~1 km} & 2000-2023 & ~ & NDVI \\ 
         & MCD12Q1.061 & Landcover IGBP scheme & & 2001-2022 & ~ & Landcover \\ 
\hline
\end{tabular}
{\raggedright *Derived from ERA5L with Eq. \ref{eq-AED}. \par}
\end{table}
```

@@

### Landcover macroclasess and validation

To analyze the LULCC, we use the IGBP scheme from the MCD12Q1 collection 6.1 from MODIS. This product has a yearly frequency from 2001 to 2022. The IGBP defines 17 classes; from these, we regrouped into ten macroclasses, as follows: classes 1-4 to forest, 5-7 to schrublands, 8-9 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands, 13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies. Thus, we have a landcover raster time-series with the ten classes for 2001 and 2023.

To validate the landcover obtained, we compare the macroclasses with the ones of a more detailed landcover map made by @Zhao2016 for Chile with samples acquired in the years 2013–2014 (LCChile). The later has a spatial resolution of 30 m and three levels of defined classes; from those, we used level 1, which fits with the macroclasses landcover. We chose the years 2013 (IGBP2013) and 2014 (IGBP2014) from landcover macrolcasses to validate with LCChile.

@@

i) resampled LCChile to the spatial resolution (500m) of the landcover macroclasses using the nearest neighbor method,
ii) took a random sample of 1000 points within continental Chile and extracted the classes that fell within each point for LCChile, IGBP2013, and IGBP2014; we considered the point extracted from LCChile as the truth and the values as the other two years as prediction
iii) calculate a confusion matrix with the classes extracted in the 1000 poitns for LCChile, IGBP2013, and IGBP2014. Calculate the performance metrics of accuracy and F1.

@@

### Landcover persistence mask 2001-2022

The time series of NDVI is affected by climatic conditions, vegetation development, seasonality, and changes in vegetation type. In this study, we want to analyze the variation in vegetation productivity in different landcover types and how it is affected by water demand, water supply, and soil moisture. In order to avoid changes due to a change in the landcover type, that will wrongly impact NDVI. We will develop a persistence mask for landcover for 2001–2023. Thereby, we reduce an important source of variation on a regional scale.

Thus, we calculated a raster mask for IGBP MODIS considering the macroclasses that remain without change for more than 80% of the years (2001–2022) per pixel, which allows us to identify the areas with no landcover change for the macroclasses.

### Landcover trend and drought indices

We calculated the surface occupied per landcover class into the five macrozones ("Norte Grande" to "Austral") per year for 2001–2023. After that, we calculated the trend's change in surface; we used the Sen' slope [@Sen1968] based on Mann-Kendall [@Kendall1975]. This way, we obtain a matrix of trends of 5 x 5 (macrozones x landcover). The aim is to later explore if the trend in landcover classes is associated with a trend in the drought indices. For this, we will use the techniques of regresion and regularization of Lasso [@Tibshirani2010] and Ridge [@Hoerl1970]. Also, we will test random forests for this purpose [@Ho1995]. We will choose the trend of landcover surface per macroclass and macrozone as the response variable and the trend of the drought indices (SPI, SPEI, EDDI, and SSI for time scales 1, 3, 6, 12, 24, and 36 months) as the predictor variables. With this analysis, we expect to gather insights regarding whether there is a pattern of climatic influence along Chile or if what is happening in Central Chile has to do with more localized climatic conditions.

@@

To estimate if there are significant positive or negative trends for the drought indices, we used the non-parametric test of Mann-Kendall [@Kendall1975]. To determine the magnitude of the trend, we used Sen's slope [@Sen1968]. Some of the advantages of applying this methodology are that the Sen's slope is not affected by outliers as regular regression does, and it is a non-paramteric method that is not affected by the distribution of the data. We applied both to the six time scales from 1981 to 2023 (monthly frequency) and the indices SPI, EDDI, SPEI, and SSI. In the case of zcNDVI (six months) was for 2000 to 2023. Thus, we have 31 trends. Also, we extracted the trend aggregated by macrozone and landcover class, obtaining a table of 31x5x5 (drought indices trends x macrozone x landcover class). We will use this data in @sec-methods_lulc to analyze if there is a strong relationship between the trends of drought indices and land cover surface within continental Chile.

### Trend in vegetation productivity without landcover change

@Vicente-Serrano2021 made a global analysis of the drought's severity trend using SPI, SPEI, and the Standardized Evapotranspiration Deficit Index (SEDI; @Vicente-Serrano2018) to evaluate AED. They indicate that the increase in hydrological drought has been due to anthropogenic effects rather than climate change. This is because the global increase in AED did not explain the change in the spatial pattern of the hydrological drought. Also, they state that _"the increase in hydrological droughts has been primarily observed in regions with high water demand and land cover change"_. We will contrast this hypothesis with what is occurring in Chile. To achieve this, we will use the landcover class type that remains more than 80% of types for 2001–2022 to evaluate the trend on zcNDVI and use this as a mask where there are low changes.

##	Impact for water supply and demand, and soil moisture in vegetation productivity within landcover types

The aim of this section is to analyze the drought indices of water demand and supply and soil moisture against vegetation to address: i) if short- or long-term time scales are most important in impacting vegetation through Chile; and ii) the strength of the correlation for the variable and the time scale. Then, we will summarize for each landcover class and macrozone. Thus, we will be able to advance in understanding how climate is affecting vegetation, considering the impact on the five macroclasses: forest, cropland, grassland, savanna, and shrubland.

@@

### Landcover macroclasess and validation

The validation of IGBP2013 and IGBP2014 with LCChile reached similar metrics of performance, having an accuracy of 0.82 and a F1 score of 0.66 (see SS1).

### Landcover persistence mask 2001-2022

The macrozones with major LULCC for 2001-2022 were "Centro", "Sur", and "Austral" with 36%, 31%, and 34%, respectively (@fig-studyArea and Table \ref{tab-landcoverTrend})

@@

@fig-LCprop shows the summary of the proportion of surface per landcover class and macrozone, derived from the mask over continental Chile.

### Landcover trend and drought indices

\begin{table}[!ht]
\caption{The value of Sen's slope trend next to the time-series plot of surface per landcover class (IGBP MCD12Q1.016) for 2001–2022 through Central Chile. Values of zero indicate that there was not a significant trend. Red dots on the plots indicate the maximum and minimum values of surface.}
\label{tab-landcoverTrend}
\includegraphics[]{../output/figs/table_var_landcover_macro.png}
\end{table}

The "Norte Chico" shows an increase in barrend land of 111 $km^2 year^{-1}$ and a reduction in the class savanna -70 $km^2 year^{-1}$. In the "Centro" and "Sur" there are changes in the Chilean matorral, with an important reduction in savanna (-136 to -318 $km^2 yr^{-1}$), and an increase in shrubland and grassland. Showing a change for more dense vegetation types. It appears to be a change from the surface occupied by cropland from the "Centro" to the "Sur." Also, there is a high increase in forest (397 $km^2 yr^{-1}$) in the "Sur," replacing the savanna lost.

@@

Regarding vegetation productivity aggregated through the macrozones in the five landcover macroclasses. In "Norte Grande" there is a increase trend of 0.02 (z-index) per decade, related with types of grassland and shrubland. There is a negative trend in "Norte Chico" with -0.04  and "Centro" with -0.02 per decade. In the "Norte Chico," savanna (-0.05) has the lowest trend, and the rest of the types are around -0.04. In "Centro," shrubland reaches -0.06, followed by grassland with -0.05, and croplands and savanna have ~-0.01 per decade. This could be associated either with a reduction in vegetation surface, a decrease in biomass, or browning [@Miranda2023]. Vegetation reached its lowest values since the year 2019, reaching an extreme condition in early 2020 and 2022 in the "Norte Chico" and Centro" (Mega Drought). The "Sur" and "Austral" show a positive trend of around 0.016 per decade (@fig-zcNDVI_var). The area most affected by a negative trend in vegetation seems to be associated with the Chilean matorral [@Fuentes2021], despite the croplands being as severely impacted by drought as native vegetation does in "Norte Chico.".

>>>>

title: Drought indices of water demand and supply, soil moisture, vegetation, and their impact on LULCC in continental Chile

@@

 Central Chile has been the focus of research studies due to the persistent decrease in water supply, which is impacting the hydrological system and vegetation development. This persistent period of water scarcity has been defined as a “Mega Drought". Our objective is to examine the effects of drought on LULCC (land use land cover change) over continental Chile using drought indices of water supply and demand, soil moisture, and vegetation productivity. For the analysis, continental Chile was divided into five zones according to a latitudinal gradient: “Norte Grande," “Norte Chico," "Centro," "Sur," and "Austral." The monthly ERA5-Land (ERA5L) variables for precipitation, temperature, and soil moisture were used. From 2001 to 2022, we used the land cover MODIS product MCD12Q1, and from 2000 to 2023, we used the NDVI (Normalized Difference Vegetation Index) product MOD13A3 collection 6.1. We estimated atmospheric evaporative demand (AED) using the Hargreaves-Samani equation with the ERA5L temperature. We used the Standardized Precipitation Index (SPI), the Standardized Precipitation Evapotranspiration Index (SPEI), the Evaporative Demand Drought Index (EDDI), the Standardized Soil Moisture Index (SSI), and the Standardized anomaly of cumulative NDVI (zcNDVI) as drought indicators. These indices were calculated for time scales of 1, 3, 6, 12, 24, and 36 months, except for zcNDVI, which was for 6 months. We analyze the trend for LULCC, vegetation productivity, and drought indices. Also, we analyzed the temporal correlation of SPI, SPEI, EDDI, and SSI with zcNDVI to gain insights into the impact of water supply and demand on vegetation productivity. Our results showed that LULCC were highest in "Centro," "Sur," and "Austral,” with 36%, 31%, and 34%, respectively. The EDDI shows that water demand has increased for all zones, with a major increase in "Norte Grande." The drought indices of water supply and soil moisture evidence a decreasing trend, which decreases at longer time scales, from "Norte Grande" to "Sur." "Austral" is the only zone that shows an increase in supply. Vegetation productivity measures by zcNDVI present a negative trend in "Norte Chico" and "Centro." Showing to be the zones most impacted by climatic conditions, the years 2020 and 2022 suffered the most extreme drought. On the other hand, forests seem to be the most resistant to drought. The types that show to be most affected by variation in climate conditions are shrublands, savannas, and croplands. The drought indices that have the capability of explaining to a major degree the variance in vegetation productivity are SSI-12, followed by SPEI-24 and SPEI-12 in "Norte Chico" and "Centro." The results indicate that “Norte Chico” and “Zona Central” are the most sensitive regions to water supply deficits longer than a year, potentially explained by a low capacity of water storage in those zones that should be further investigated.

@@

  docx:

@@

The sixth assessment report (AR6) of the IPCC [@IPCC2023] indicates that human-induced greenhouse gas emissions have increased the frequency and/or intensity of some weather and climate extremes, and the evidence has been strengthened since AR5 [@IPCC2013]. There is a high degree of confidence that rising temperatures will increase the land area where droughts will occur more frequently and with greater severity [@IPCCCH112021]. Furthermore, drought increases tree mortality and triggers changes in land cover and, consequently, land use, thus impacting ecosystems [@Crausbay2017]. Nevertheless, there is a lack of understanding of how the alteration in water supply and demand is affecting land cover transformations.

The primary cause of drought is precipitation, and temperature makes it worse [@Luo2017]. Drought impacts soil moisture, hydrological regimes, and vegetation productivity. Initially, drought was commonly classified as meteorological, hydrological, and agricultural [@Wilhite1985]. Lately, @Loon2016 and @AghaKouchak2021 have given an updated definition of drought for the Anthropocene, suggesting that it should be considered the feedback of humans' decisions and activities that drives the anthropogenic drought. Even though it has been argued that those definitions do not fully address the ecological dimensions of drought, @Crausbay2017 proposed the ecological drought definition as "an episodic deficit in water availability that drives ecosystems beyond thresholds of vulnerability, impacts ecosystem services, and triggers feedback in natural and/or human systems.". Moreover, many ecological studies have misinterpreted how to characterize drought, for example, sometimes considering "dry" conditions as "drought" [@Slette2019]. On the other hand, the AR6 [@IPCC2023] predicts that many regions of the world will experience more severe agricultural and ecological droughts even if global warming stabilizes at 1.5°–2°C. Then, there is a challenge in conducting drought research, especially to evaluate its impact on ecosystems.

@@

For the spatiotemporal assessment of drought impact (i.e., by water supply and demand) on land cover changes, we need climatic realiable variables such as precipitation, temperature, soil moisture, land cover, and vegetation status. For developing countries like Chile, the weather networks present several disadvantages, such as gaps, a short history, and low-quality data. Reanalysis data, as the ERA5-Land (ERA5L) [@MunozSabater2021] provides hourly climatic information (precipitation, temperature, and soil moisture) without gaps since 1950 with global extension. ERA5L has already been used for drought assessment using the Standardized Precipitation-Evapotranspiration Index (SPEI) [@Nouri2023] and for flash drought [@Wang2023] by analyzing soil moisture and evapotranspiration. On the other hand, satellite remote sensing [@West2019; @AghaKouchak2015] is the primary method to evaluate how drought impacts vegetation dynamics. The Moderate-Resolution Imaging Spectroradiometer (MODIS) can be used to get vegetation drought indices (VDI), which are often used as proxies for productivity [@Paruelo2016; @Schucknecht2017]. Besides, land use and land cover (LULC) change can be driven by drought [@Tran2019; @Akinyemi2021]. To analyze these changes, multiple LULC products exist [@Grekousis2015]. One of those that provides time series since 2001 is the MCD12Q1 [@Friedl2019] from MODIS. The variation in water supply and demand is finally reflected in the total water storage (TWS). The Gravity Recovery and Climate Experiment (GRACE), which allows analyzing changes in water availability at coarse resolution, can retrieve the TWS [@Ahmed2014; @Ma2017]. We can find drought indices of supply (i.e., precipitation) and demand (i.e., temperature) using climatic reanalysis (ERA5L) and vegetation data (MODIS). This lets us figure out how drought changes LULC. Further, the TWS can be assessed with regard to the changes in water supply and demand to gain insight into the impact on water storage.

@@

In this research, we aim to analyze the impact of drought on different types of ecosystems (land cover classes) in continental Chile. Our specific goals are: i) to analyze the trend on multi-scalar drought indices for water demand and supply, soil moisture, and vegetation productivity for 1981–2023/2001–2023; ii) to assess the LULC change for 2001–2021 and how it relates to drought indices; iii) to evaluate the relationship between zcNDVI, a measure of vegetation productivity, and drought indices for water demand and supply and soil moisture; and iv) to assess if the observed changes in the drought indices are linked to TWS.

@@

Continetal Chile has a diverse climate condition from north to south and east to west [@Aceituno2021] (@fig-studyArea), which determines its great ecosystem diversity (@fig-LCprop). The Andes Mountains are a main factor in latitudinal variation [@Garreaud2009]. To describe the climate and ecosystem of Chile, we use the Koppen-Geiger release by @Beck2023 and the land cover type persistance of 80% of times for 2001–2022, from the IGBP classification scheme [@Friedl2019] (see @sec-methods_lulc). "Norte Grande" and "Norte Chico" predominate in an arid desert climate with hot (Bwh) and cold (Bwk) temperatures. At the south of "Norte Chico," the climate changes to an arid steppe with cold temperatures (Bsk). Mainly, the land is barren, with a minor surface of vegetation types such as shrubland and grassland. In the zones "Centro" and the north half of "Sur," the main climate is Mediterranean, with warmer to hot summers (Csa and Csb). There is a significant amount (50%) of Chilean matorral (shrubland and savanna, [@Fuentes2021]), then grassland (16%), forest (8%), and croplands (5%), in "Centro." The south part of "Sur" and the north part of "Austral" are dominated by an oceanic climate (Cfb). Those zones are high in forest and grassland. The southern part of the country has a tundra climate, and in Patagonia, it is a cold semi-arid area with an extended surface of grassland, forest, and, to a lesser extent, savanna.

@@

```{=latex}
\begin{table}[!ht]
\caption{Description of the earth observation data used }
\label{tab-desEOD}
\small
\centering
\begin{tabular}{p{0.13\textwidth}cp{0.3\textwidth}p{0.095\textwidth}ccc}
\hline
\multirow{1}{*}{\centering Product} & Sub-product & Variable & Spatial Resolution  & Period & Units & Short Name \\ 
\hline
\multirow{4}{*}{ERA5L} & ~ & Precipitation & \multirow{4}{*}{~0.1°} & \multirow{4}{*}{1981-2023} & mm & P \\ 
         &  & Maximum temperature & ~ & & $°C$ & $T_{max}$ \\ 
         &  & Minimum temperature & ~ & & $°C$ & $T_{min}$ \\ 
         &  & Volumetric Soil Water Content at 1m & ~ & & $m3/m3$ & SM \\ 
ERA5L* & & Atmospheric Evaporative Demand & 0.1° & 1981-2023 & mm & AED \\
        \multirow{2}{*}{MODIS} & MOD13A3.061 & Normalized Difference Vegetation Index & \multirow{2}{*}{~1 km} & 2000-2023 & ~ & NDVI \\ 
         & MCD12Q1.061 & land cover IGBP scheme & & 2001-2022 & ~ & land cover \\ 
\hline
\end{tabular}
{\raggedright *Derived from ERA5L with Eq. \ref{eq-AED}. \par}
\end{table}
```

@@

### land cover macroclasess and validation

To analyze the LULCC, we use the IGBP scheme from the MCD12Q1 collection 6.1 from MODIS. This product has a yearly frequency from 2001 to 2022. The IGBP defines 17 classes; from these, we regrouped into ten macroclasses, as follows: classes 1-4 to forest, 5-7 to schrublands, 8-9 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands, 13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies. Thus, we have a land cover raster time-series with the ten classes for 2001 and 2023.

To validate the land cover obtained, we compare the macroclasses with the ones of a more detailed land cover map made by @Zhao2016 for Chile with samples acquired in the years 2013–2014 (LCChile). The later has a spatial resolution of 30 m and three levels of defined classes; from those, we used level 1, which fits with the macroclasses land cover. We chose the years 2013 (IGBP2013) and 2014 (IGBP2014) from land cover macrolcasses to validate with LCChile.

@@

i) resampled LCChile to the spatial resolution (500m) of the land cover macroclasses using the nearest neighbor method,
ii) took a random sample of 1000 points within continental Chile and extracted the classes that fell within each point for LCChile, IGBP2013, and IGBP2014; we considered the point extracted from LCChile as the truth and the values as the other two years as prediction
iii) calculate a confusion matrix with the classes extracted in the 1000 poitns for LCChile, IGBP2013, and IGBP2014. Calculate the performance metrics of accuracy and F1.

@@

### land cover persistence mask 2001-2022

The time series of NDVI is affected by climatic conditions, vegetation development, seasonality, and changes in vegetation type. In this study, we want to analyze the variation in vegetation productivity in different land cover types and how it is affected by water demand, water supply, and soil moisture. In order to avoid changes due to a change in the land cover type, that will wrongly impact NDVI. We will develop a persistence mask for land cover for 2001–2023. Thereby, we reduce an important source of variation on a regional scale.

Thus, we calculated a raster mask for IGBP MODIS considering the macroclasses that remain without change for more than 80% of the years (2001–2022) per pixel, which allows us to identify the areas with no land cover change for the macroclasses.

### land cover trend and drought indices

We calculated the surface occupied per land cover class into the five macrozones ("Norte Grande" to "Austral") per year for 2001–2023. After that, we calculated the trend's change in surface; we used the Sen' slope [@Sen1968] based on Mann-Kendall [@Kendall1975]. This way, we obtain a matrix of trends of 5 x 5 (macrozones x land cover). The aim is to later explore if the trend in land cover classes is associated with a trend in the drought indices. For this, we will use the techniques of regresion and regularization of Lasso [@Tibshirani2010] and Ridge [@Hoerl1970]. Also, we will test random forests for this purpose [@Ho1995]. We will choose the trend of land cover surface per macroclass and macrozone as the response variable and the trend of the drought indices (SPI, SPEI, EDDI, and SSI for time scales 1, 3, 6, 12, 24, and 36 months) as the predictor variables. With this analysis, we expect to gather insights regarding whether there is a pattern of climatic influence along Chile or if what is happening in Central Chile has to do with more localized climatic conditions.

@@

To estimate if there are significant positive or negative trends for the drought indices, we used the non-parametric test of Mann-Kendall [@Kendall1975]. To determine the magnitude of the trend, we used Sen's slope [@Sen1968]. Some of the advantages of applying this methodology are that the Sen's slope is not affected by outliers as regular regression does, and it is a non-paramteric method that is not affected by the distribution of the data. We applied both to the six time scales from 1981 to 2023 (monthly frequency) and the indices SPI, EDDI, SPEI, and SSI. In the case of zcNDVI (six months) was for 2000 to 2023. Thus, we have 31 trends. Also, we extracted the trend aggregated by macrozone and land cover class, obtaining a table of 31x5x5 (drought indices trends x macrozone x land cover class). We will use this data in @sec-methods_lulc to analyze if there is a strong relationship between the trends of drought indices and land cover surface within continental Chile.

### Trend in vegetation productivity without land cover change

@Vicente-Serrano2021 made a global analysis of the drought's severity trend using SPI, SPEI, and the Standardized Evapotranspiration Deficit Index (SEDI; @Vicente-Serrano2018) to evaluate AED. They indicate that the increase in hydrological drought has been due to anthropogenic effects rather than climate change. This is because the global increase in AED did not explain the change in the spatial pattern of the hydrological drought. Also, they state that _"the increase in hydrological droughts has been primarily observed in regions with high water demand and land cover change"_. We will contrast this hypothesis with what is occurring in Chile. To achieve this, we will use the land cover class type that remains more than 80% of types for 2001–2022 to evaluate the trend on zcNDVI and use this as a mask where there are low changes.

##	Impact for water supply and demand, and soil moisture in vegetation productivity within land cover types

We analyze the drought indices of water demand and supply and soil moisture against vegetation to address: i) if short- or long-term time scales are most important in impacting vegetation through Chile; and ii) the strength of the correlation for the variable and the time scale. Then, we will summarize for each land cover class and macrozone. Thus, we will be able to advance in understanding how climate is affecting vegetation, considering the impact on the five macroclasses having vehetation: forest, cropland, grassland, savanna, and shrubland.

@@

### land cover macroclasess and validation

For vegetation, we obtained and use hereafter five macroclasses of land cover from IGBP MODIS: forest, shrubland, savanna, grassland, and croplands. @fig-studyArea c shows the spatial distribution of the macroclasses through Chile for the year 2022. The validation of IGBP2013 and IGBP2014 with LCChile reached near the same metrics of performance, having an accuracy of ~0.82 and a F1 score of ~0.66 (see SS1).

### land cover persistence mask 2001-2022

@fig-studyArea d, shows the macroclasses of land cover persistance (80%) during 2021-2022, respectively. Within continental Chile, forest is the vegetation type with highest surface with 135,00 $km^2$, followed by grassland (73,176 $km^2$), savanna (54,410 $km^2$), shrubland (24,959 $km^2$), and cropland (3,100 $km^2$) (\ref{tab-land coverSurf}).  The macrozones with major LULCC for 2001-2022 were "Centro", "Sur", and "Austral" with 36%, 31%, and 34%, respectively (@fig-studyArea and Table \ref{tab-land coverTrend}); of its surface that changes the type of land cover. @fig-LCprop shows the summary of the proportion of surface per land cover class and macrozone, derived from the persistance mask over continental Chile.

```{=latex}
\begin{table}[!ht]
\caption{Surface of the land cover class that persist during 2001-2022}
\label{tab-land coverSurf}
\includegraphics[width = .5\textwidth]{../output/figs/table_surface_landcover_macrozone.png}
\end{table}
```

@@

### land cover trend and drought indices

\begin{table}[!ht]
\caption{The value of Sen's slope trend next to the time-series plot of surface per land cover class (IGBP MCD12Q1.016) for 2001–2022 through Central Chile. Values of zero indicate that there was not a significant trend. Red dots on the plots indicate the maximum and minimum values of surface.}
\label{tab-land coverTrend}
\includegraphics[]{../output/figs/table_var_landcover_macro.png}
\end{table}

The "Norte Chico" shows an increase in barrend land of 111 $km^2 year^{-1}$ and a reduction in the class savanna of 70 $km^2 year^{-1}$. In the "Centro" and "Sur," there are changes in the Chilean matorral, with an important reduction in savanna (136 to 318 $km^2 yr^{-1}$), and an increase in shrubland and grassland. Showing a change for more dense vegetation types. It appears to be a shift in the area of cropland from the "Centro" to the "Sur." Also, there is a high increase in forest (397 $km^2 yr^{-1}$) in the "Sur," replacing the savanna lost.

Further, we want to address whether the trend in land cover change for 2001–2023 is associated with trends in drought indices of water demand and supply and/or soil moisture for macrozone and land cover macroclasses. From the three methods tested, Ridge, Lasso, and Random Forest, neither gives significant results regarding whether the trend in a drought index for any time scale explains the trend in land cover change. Nevertheless, in "Norte Chico" and "Centro," there is a decrease in croplands and savanna and an increase in barren land, which is associated with the variation in drought indices. Mainly for a decrease in water supply (SPI and SSI) and an increase in water demand (EDDI). However, due to the high variability from north to south in Chile, the climatic condition (arid, semi-arid, and humid), and the land cover type, we believe that only in those zones could the LULCC be driven to some degree by drought.

@@

```{r fig-zcNDVI_var}
#| echo: false
#| fig-pos: "!ht"
#| fig-cap: "(a) Map of the linear trend of the index zcNDVI-6 for 2001–2023. Greener colors indicate a positive trend; reder colors correspond to a negative trend and a decrease in vegetation productivity. Grey colors indicate either no vegetation or a change in land cover type for 2001–2022. (b) Temporal variation of zcNDVI-6 aggregated at macrozone level within continental Chile. Each horizontal panel corresponds to a macrozone from 'Norte Grande' to 'Austral'."
#| out-width: "\\textwidth"
knitr::include_graphics("../output/figs/temporal_variation_zcNDVI6_macrozonas_con_mapa.png")
```
